---
title: "Cytotoxic immune infiltration effects on breast cancer survival and immunotherapy response"
author: "cyy36@cam.ac.uk"
date: today
bibliography: references.bib
csl: nature.csl
link-citations: true
execute:
  warning: false
  message: false
  echo: false
  cache: false
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    embed-resources: true
    theme:
      light: flatly
      dark: darkly
---

# Cover Page {.unnumbered .unlisted}

<center>
![](https://onlinecareeraccelerators.pace.cam.ac.uk/hubfs/CAM%20DS%20PACE/Professional%20and%20Continuing%20Education_V_reversed%20colour%201.svg)
</center>

### MSt Healthcare Data Science {.unnumbered .unlisted}

#### Authentication of practice

|   |   |
|---|---|
|I confirm that I have fully read and understood the [assignment brief](https://vle.pace.cam.ac.uk/mod/folder/view.php?id=2440175) for this module.| **Y** |

#### Details

|   |   |
|---|---|
|Name: Chung Yan|Surname: Yu|
|Submission Date| 10 April 2026 |
|Word Count: whole assignment including codes|  |
|Word Count: main body excluding abstract, references and supplementary materials|  |

#### Permission to share your assignment

I do give permission to share my assignment with future MSt participants.

#### University statement of originality

This assignment is the result of my own work and includes nothing which is the outcome of work done in collaboration except as declared in the Preface and specified in the text. It is not substantially the same as any that I have previously submitted for a degree or diploma or other qualification at the University of Cambridge or any other university or similar institution, or that is being concurrently submitted, except as declared in the Preface and specific in the text. I further state that no substantial part of my Portfolio has already been submitted, nor is being concurrently submitted for any such degree, diploma or other qualification at the University of Cambridge or any other university or similar institution except as declared in the Preface and specified in the text.

|   |   |
|---|---|
|I confirm the statement of originality as above| **Y** |


#### Questions for reflection

Self-assessment is an important aspect of feedback literacy, which is, in turn, key to the development of expertise. As you proceed through the MSt Healthcare data science programme, we hope that you will make use of the following prompts to assess your own work on assignments. Specific assignment briefs will likely indicate which of these to address for which assessments, but, in general, we expect you to respond to one or two for each assignment on your course.

For each of the questions, do not spend too long answering – keep it brief. For each question you answer, limit yourself to no more than three items. And please remember, this is optional and developmental: these cover sheets are designed to create space for self-assessment and feedback dialogue, rather than additional assignment workload.

1. Which aspects of this assignment are you most uncertain about and/or would most like to receive feedback on?

- A1 size is more of a newer size/scale for me, so I might have packed things too densely here. To keep the captions light, I also stuffed quite a lot of stats in the plots itself, not sure if that is done well or is suitable.

2. What elements are you left pondering after this assignment that you would like to discuss further?

- It feels a bit much to add an addendum to this write up, so maybe I will just write things out here. The smaller size of A1 definitely threw me off a bit, as I really struggled to fit everything into the poster. I might also have spent a bit too much time playing around with the Cox assumption testing rather than ggplot itself, which feels like a shame. I do hope I will still have opportunities to really give ggplot a good go in future assignments. Overall this assignment felt tougher than the previous qmd based tasks, possibly due to the constant trial and error with the layout. With the qmd, everything just scrolls down sequentially so there is less mental load there. I had thought this assignment would be the lightest one so far, given my previous experience with academic posters, but it ended up being the opposite. Really trying to make everything fit in just right takes a lot of effort, so much so that I think some of my previous thoughts/comments on other already published visualisations we discussed in class too harsh. 

3. How have you incorporated feedback from peers and tutors into this assignment?

- Managed to try the Cox assumption testing this time. I think some of the assumptions might be borderline violated, but might still work given the circumstance?

4. How, and to what extent, have you been able to incorporate feedback on previous course work into this assignment?


5. Using the wording in the rubric, how would you describe the quality of the different aspects of your work?


### Declaration of the use of generative AI {.unnumbered .unlisted}

|   |   |
|---|---|
|Which permitted use of generative AI are you acknowledging?| Semantic search of literature and notes, outline creation, output formatting, informed feedback, code generation |
|Which generative AI tool did you use (name and version)?| Claude Code v2.1.1x (Opus 4.6), M365 CoPilot, GitHub CoPilot |
|What did you use the tool for?| Searching for relevant journal publications, collating notes in my Obsidian vault, `.bib` file curation, code debugging+generation |
|How have you used or changed the generative AI's output| My collated notes always stay in point form so that I write out the paragraphs in my own words. Feedback provided by the GenAI models are weighed and assessed before being acted on. |


> The GitHub repository of this project is found [here](https://github.com/joelyu/HDS-W4.git). The poster PDF is found [here](https://joelyu.github.io/HDS-W4/poster/poster.pdf)

# Design rationale

This poster targets researchers at cancer, immuno-oncology or genomic research conferences, where readers are already fluent in survival analysis, cancer hallmarks and subtypes, and scientific poster conventions in general. 

The poster is intended to be read top-to-bottom, with the main content focused on the left, and more supplementary materials such as illustrations and methods to the right. The headline numbers on the top-left anchor the reader with the top strip, and along with the background statements, orient the reader with the premise. Clear section headings with consistent colouring guide the reader down through figures 1 and 2. The conclusion sits at the bottom right as the natural end of the poster, with the main message used as the header for a quick and easy read.

The Okabe-Ito colour palette is used for colour-blind accessibility, and each cohort is consistently associated with one colour. The orange colour makes up the colour theme of the poster as used in banners and cell illustrations.

Minimalistic plots are used to put the main focus on the data, to the extent some panels share a y-axis. Relevant statistics are embedded with the legend keys rather than buried in the captions. Data provenance and processing are denoted in visual flow charts for intuitive comprehension. Obvious conventions such as p-value significance notation are presumed self-explanatory given the background of the audience. The Source Sans font was used for a modern finish, save for a monospaced font used for R package names. 


# Figures & code

## Setup

### Data analysis

Analyses are done via scripts in the repository, plot generation is done in code chunks in this document.

```{r}
#| label: setup
#| include: false
#| code-summary: "Setup code"

# Only load packages used directly in this document's chunks.
# Data-pipeline packages (dplyr, tidyr, tibble, survival, GEOquery,
# cBioPortalData) are loaded by the sourced scripts below and become
# available here after source().
suppressPackageStartupMessages({
  library(ggplot2)    # all figure chunks
  library(survminer)  # ggsurvplot in fig-2a-km
})

# Source the full analysis pipeline
# The scripts handle their own caching via CSV files in data/processed/,
# so re-running is fast on subsequent renders.
invisible(capture.output(suppressMessages({
  source("scripts/00-fetch-data.R")
  source("scripts/01-immune-scoring.R")
  source("scripts/02-explore.R")
  source("scripts/03-pca-cytotoxic.R")
  source("scripts/04-confirmatory-cox.R")
  source("scripts/05-ispy2.R")
})))
```

### Plot theme and palette

Setting up usage of Okabe-Ito colourblind-safe palette and a clean and minmal theme. Assigning colors to cohorts as well.

```{r}
#| label: plot-theme
#| echo: true
#| code-fold: true
#| code-summary: "Theme setup code"

# ------------------------------------------------------------------
# Palette — Okabe-Ito colourblind-safe set, assigned to narrative
# roles so each colour has one job across all figures.
#
#   METABRIC (discovery cohort)         → orange/Okabe-Ito vermillion,
#   TCGA (validation cohort)            → blue/Okabe-Ito blue, 
#   Pembrolizumab arm (Fig 3B only)     → purple/Okabe-Ito reddish purple,
#   Control / neutral / background      → dark grey
#
# ------------------------------------------------------------------

PALETTE <- list(
  metabric     = "#D55E00",  # Okabe-Ito vermillion
  tcga         = "#0072B2",  # Okabe-Ito blue
  pembro       = "#CC79A7",  # Okabe-Ito reddish purple
  control      = "#4D4D4D",  # dark grey
  neutral      = "#4D4D4D"   # alias
)

# Named vectors for ggplot scale_* helpers
COHORT_COLOURS <- c(
  METABRIC = PALETTE$metabric,
  TCGA     = PALETTE$tcga
)

ARM_COLOURS <- c(
  Pembrolizumab = PALETTE$pembro,
  Control       = PALETTE$control
)

# ------------------------------------------------------------------
# theme_hds() — clean, Tufte-inspired, axis-only, white background
# ------------------------------------------------------------------
theme_hds <- function(base_size = 14) {
  theme_classic(base_size = base_size) +
    theme(
      # Titles and axes
      plot.title        = element_text(face = "bold", size = base_size * 1.15),
      plot.subtitle     = element_text(colour = "grey40", size = base_size * 0.95),
      axis.title        = element_text(face = "plain", size = base_size),
      axis.text         = element_text(size = base_size * 0.9),
      axis.line         = element_line(colour = "black", linewidth = 0.4),
      axis.ticks        = element_line(colour = "black", linewidth = 0.4),
      # Legend
      legend.title      = element_text(face = "bold", size = base_size * 0.95),
      legend.text       = element_text(size = base_size * 0.9),
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.position   = "right",
      # Panels and strip
      panel.background  = element_rect(fill = "white", colour = NA),
      panel.grid        = element_blank(),
      strip.background  = element_blank(),
      strip.text        = element_text(face = "bold", size = base_size * 0.95),
      # Overall
      plot.background   = element_rect(fill = "white", colour = NA)
    )
}

# Convenience scale functions — consistent cohort encoding
scale_colour_cohort <- function(...) scale_colour_manual(values = COHORT_COLOURS, ...)
scale_fill_cohort   <- function(...) scale_fill_manual(values = COHORT_COLOURS, ...)
scale_colour_arm    <- function(...) scale_colour_manual(values = ARM_COLOURS, ...)
scale_fill_arm      <- function(...) scale_fill_manual(values = ARM_COLOURS, ...)

# Diverging gradient for heatmaps: blue ↔ white ↔ vermillion
# (reuses METABRIC and TCGA colours as palette anchors)
scale_fill_diverging <- function(limits = c(-1, 1), ...) {
  scale_fill_gradient2(
    low      = PALETTE$tcga,
    mid      = "white",
    high     = PALETTE$metabric,
    midpoint = 0,
    limits   = limits,
    ...
  )
}
```

## Figures

### Figure 1 — Identifying representative immune population across two cohorts

#### 1A. Forest plot

14 cell populations, PAM50-adjusted HRs, fully-adjusted HRs for cytotoxic cells, both cohorts (METABRIC, TCGA)

```{r}
#| label: plot-1a-forest
#| echo: true
#| code-fold: true
#| code-summary: "Figure 1A code"
#| fig-width: 9
#| fig-height: 8

# --- Prepare data -----------------------------------------------------------
# `comparison` comes from 02-explore.R (long format, both models)
# Saved to data/processed/cox_comparison.csv but the in-memory name is `comparison`
forest_data <- comparison %>%
  filter(model == "PAM50-adjusted") %>%
  mutate(
    cohort = factor(cohort, levels = c("metabric", "tcga"),
                    labels = c("METABRIC", "TCGA"))
  )

# Cell type order: by METABRIC PC1 loading (from script 03, loadings_m)
# This ties Fig 1A visual ordering to Fig 2A (loading plot) — same sequence
cell_order <- names(sort(abs(loadings_m), decreasing = FALSE))  # bottom-up for ggplot
forest_data <- forest_data %>%
  mutate(cell_type = factor(cell_type, levels = cell_order))

# --- Append the fully-adjusted cytotoxic row (f3/g3 PRIMARY models) ---------
# These come from cox_robustness in 04-confirmatory-cox.R. Adding them as a
# new top row lets the forest plot show both the discovery screen (PAM50-
# adjusted, 14 cell types) AND the confirmatory survival result for cytotoxic
# (fully adjusted for PAM50 + age + stage) in one figure. The top row will be
# highlighted in Illustrator to signal its different adjustment level.
primary_row <- cox_robustness %>%
  filter(grepl("PRIMARY", model)) %>%
  transmute(
    cohort    = factor(cohort, levels = c("METABRIC", "TCGA")),
    cell_type = "Cytotoxic (fully adjusted)",
    hr        = HR,
    hr_lower  = HR_lower,
    hr_upper  = HR_upper,
    # Single pre-specified confirmatory test — use raw p, no FDR correction
    fdr       = p_value
  )

# Extend the cell-type ordering so the new row sits at the very top
cell_order_ext <- c(cell_order, "Cytotoxic (fully adjusted)")

# Tag analysis stage for shape encoding (● screening, ◆ confirmatory)
forest_data <- bind_rows(
  forest_data %>%
    select(cohort, cell_type, hr, hr_lower, hr_upper, fdr) %>%
    mutate(analysis = "Screening"),
  primary_row %>% mutate(analysis = "Confirmatory")
) %>%
  mutate(
    cell_type = factor(cell_type, levels = cell_order_ext),
    analysis  = factor(analysis, levels = c("Screening", "Confirmatory")),
    sig_label = dplyr::case_when(
      fdr < 0.001 ~ "***",
      fdr < 0.01  ~ "**",
      fdr < 0.05  ~ "*",
      TRUE        ~ ""
    )
  )

# --- Build legend n counts from cox_robustness ------------------------------
# Screening n = Unadjusted model (full PAM50-ready subset)
# Confirmatory n = PRIMARY model (stage-complete subset)
get_n <- function(coh, pattern) {
  cox_robustness %>%
    filter(cohort == coh, grepl(pattern, model)) %>%
    pull(n) %>% dplyr::first()
}
n_m_scr <- get_n("METABRIC", "^Unadjusted$")
n_m_cnf <- get_n("METABRIC", "PRIMARY")
n_t_scr <- get_n("TCGA", "^Unadjusted$")
n_t_cnf <- get_n("TCGA", "PRIMARY")

cohort_labels <- c(
  "METABRIC" = sprintf("METABRIC (n = %s / %s)",
                       format(n_m_scr, big.mark = ","),
                       format(n_m_cnf, big.mark = ",")),
  "TCGA"     = sprintf("TCGA (n = %s / %s)",
                       format(n_t_scr, big.mark = ","),
                       format(n_t_cnf, big.mark = ","))
)

# Dodge width for per-row cohort offset
dodge_width <- 0.5

# --- Build plot -------------------------------------------------------------
fig_1a <- ggplot(forest_data,
                 aes(x = hr, y = cell_type, colour = cohort, shape = analysis)) +
  # Reference line at HR = 1.0
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = PALETTE$neutral, linewidth = 0.4) +
  # 95% CI error bars
  geom_errorbarh(
    aes(xmin = hr_lower, xmax = hr_upper),
    height = 0,
    position = position_dodge(width = dodge_width),
    linewidth = 0.6
  ) +
  # HR point estimates (shape distinguishes screening vs confirmatory)
  geom_point(
    size = 3.5,
    position = position_dodge(width = dodge_width)
  ) +
  # FDR significance asterisks, placed just right of each CI upper bound
  geom_text(
    aes(x = hr_upper, label = sig_label),
    position = position_dodge(width = dodge_width),
    hjust = -0.2, vjust = 0.5, size = 4, show.legend = FALSE
  ) +
  # Log-scale x-axis (HR is multiplicative)
  scale_x_continuous(
    trans = "log10",
    breaks = c(0.6, 0.7, 0.8, 0.9, 1.0, 1.1),
    labels = c("0.6", "0.7", "0.8", "0.9", "1.0", "1.1")
  ) +
  # Clean cell type labels: replace underscores with spaces
  scale_y_discrete(labels = function(x) gsub("_", " ", x)) +
  scale_colour_cohort(
    labels = cohort_labels,
    guide  = guide_legend(override.aes = list(shape = 16))
  ) +
  scale_shape_manual(
    values = c("Screening" = 16, "Confirmatory" = 18),
    guide  = "none"
  ) +
  labs(
    x = "Hazard ratio (per 1 SD, 95% CI)",
    y = "Immune cell populations",
    colour = NULL,
    shape  = NULL
  ) +
  theme_hds() +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.box = "vertical"
  )

# Save the plot as SVG, fiddle with the size here.
ggsave("poster/fig_1a.svg", fig_1a, width = 7, height = 8, device = "svg")
fig_1a
```

#### 1B. Lollipop/Cleveland plot

Comparing PC1 loadings across METABRIC and TCGA cohorts.

```{r}
#| label: plot-1b-loadings
#| echo: true
#| code-fold: true
#| code-summary: "Figure 1B code"
#| fig-width: 7
#| fig-height: 6

# --- Cleveland dot plot: PC1 loading per cell type, both cohorts -----------
# loadings_m and loadings_t come from 03-pca-cytotoxic.R
# Use absolute values so the comparison is on magnitude (direction is
# arbitrary in PCA — both cohorts already flipped to positive PC1 = hot)
loadings_clev <- data.frame(
  cell_type = names(loadings_m),
  metabric  = abs(loadings_m),
  tcga      = abs(loadings_t[names(loadings_m)])
) %>%
  # Order top-down by METABRIC loading (cytotoxic at the top)
  mutate(
    cell_type = factor(
      cell_type,
      levels = cell_type[order(metabric)]  # ascending → ggplot puts largest at top
    )
  )

# Long format for the dots
loadings_clev_long <- loadings_clev %>%
  tidyr::pivot_longer(
    cols = c(metabric, tcga),
    names_to = "cohort",
    values_to = "loading"
  ) %>%
  mutate(
    cohort = factor(cohort,
                    levels = c("metabric", "tcga"),
                    labels = c("METABRIC", "TCGA"))
  )

cell_order_clev <- levels(loadings_clev$cell_type)  # used by Fig 1C for shared y-ordering

# Extract PC1 variance explained for the subtitle (computed once, data-driven)
pc1_var_m <- summary(pca_m)$importance[2, 1] * 100
pc1_var_t <- summary(pca_t)$importance[2, 1] * 100

fig_1b <- ggplot(loadings_clev_long, aes(x = loading, y = cell_type)) +
  # Connecting line per cell type — its length shows cross-cohort disagreement
  geom_line(aes(group = cell_type), colour = "grey70", linewidth = 0.6) +
  # Cohort dots
  geom_point(aes(colour = cohort), size = 3.5) +
  scale_y_discrete(labels = function(x) gsub("_", " ", x)) +
  scale_colour_cohort() +
  labs(
    x = "Absolute PC1 loading",
    y = "Immune cell populations",
    colour = NULL,
    subtitle = sprintf("PC1 explains %.1f%% (METABRIC), %.1f%% (TCGA) of variance",
                       pc1_var_m, pc1_var_t)
  ) +
  theme_hds() +
  theme(
    legend.position = "top",
    legend.justification = "left"
  )

# Save the plot as SVG, fiddle with the size here.
ggsave("poster/fig_1b.svg", fig_1b, width = 5, height = 6, device = "svg")
fig_1b
```

#### 1C. Collinearity heatmap

Examining collinearity between cell type scores in METABRIC.

```{r}
#| label: plot-1c-heatmap
#| echo: true
#| code-fold: true
#| code-summary: "Figure 1C code"
#| fig-width: 7
#| fig-height: 6

# --- METABRIC pairwise correlation between cell type scores --------------
score_cols_2b <- setdiff(
  colnames(scores_metabric),
  c("patient_id", "cohort", "T_cells")
)

corr_m <- cor(surv_metabric[, score_cols_2b], use = "pairwise.complete.obs")

# Reorder the matrix rows/columns to match the plotting order (Fig 1B Cleveland)
# BEFORE blanking the triangle, so the blanked half aligns with the visual layout
corr_m <- corr_m[cell_order_clev, cell_order_clev]

# Blank the LOWER matrix triangle — with ascending factor ordering
# (cytotoxic at the top of the y-axis), lower.tri() blanks the top-right
# visually, leaving the bottom-left triangle visible. Cytotoxic sits at the
# top with the longest visible row extending rightward from the diagonal.
corr_m[upper.tri(corr_m)] <- NA

# Long format — factor levels match the matrix ordering
corr_long <- as.data.frame(corr_m) %>%
  tibble::rownames_to_column("cell_a") %>%
  tidyr::pivot_longer(-cell_a, names_to = "cell_b", values_to = "r") %>%
  filter(!is.na(r)) %>%  # drop the blanked lower triangle
  mutate(
    cell_a = factor(cell_a, levels = rev(cell_order_clev)),
    cell_b = factor(cell_b, levels = cell_order_clev)
  )

fig_1c <- ggplot(corr_long, aes(x = cell_a, y = cell_b, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(
    aes(label = sprintf("%.2f", r)),
    size = 2.5,
    colour = ifelse(abs(corr_long$r) > 0.6, "white", "black")
  ) +
  scale_fill_diverging(limits = c(-1, 1), name = "Pearson r") +
  scale_x_discrete(labels = function(x) gsub("_", " ", x)) +
  scale_y_discrete(labels = function(x) gsub("_", " ", x)) +
  labs(
    x = "Immune cell populations",
    y = NULL
  ) +
  theme_hds() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.border = element_rect(colour = NA, fill = NA)
  )

# Save the plot as SVG, fiddle with the size here.
ggsave("poster/fig_1c.svg", fig_1c, width = 7, height = 6, device = "svg")
fig_1c
```

**Figure 1** | **Identification of cytotoxic cells as a cross-cohort immune axis in breast cancer.**
**(A)** PAM50-adjusted hazard ratios for 14 immune populations in METABRIC (n = 1,756) and TCGA-BRCA (n = 981); top row shows the fully adjusted cytotoxic result. Error bars: 95% CI; \* \*\* \*\*\* denote FDR < 0.05, 0.01, 0.001.
**(B)** Absolute PC1 loadings per population across both cohorts.
**(C)** Pairwise Pearson correlations in METABRIC, ordered by PC1 loading.

### Figure 2 — Association with survival and response

#### 2A. Kaplan-Meier curve

Median split of cytotoxic scores in the METABRIC cohort, unadjusted.

```{r}
#| label: plot-2a-km
#| echo: true
#| code-fold: true
#| code-summary: "Figure 2A code"
#| fig-width: 7
#| fig-height: 5

# --- Median split on cytotoxic score, METABRIC ---------------------------
surv_m_km <- surv_metabric %>%
  mutate(
    cyto_group = factor(
      ifelse(Cytotoxic_cells >= median(Cytotoxic_cells),
             "High cytotoxic", "Low cytotoxic"),
      levels = c("Low cytotoxic", "High cytotoxic")
    )
  )

fit_km_m <- survfit(Surv(os_months, os_event) ~ cyto_group, data = surv_m_km)

# survminer::ggsurvplot returns a list with $plot — extract it for theming
km_plot <- ggsurvplot(
  fit_km_m,
  data         = surv_m_km,
  conf.int     = TRUE,
  palette      = c(PALETTE$neutral, PALETTE$metabric),
  pval         = TRUE,
  pval.coord   = c(150, 0.85),
  xlab         = "Time (months)",
  ylab         = "Overall survival probability",
  legend.title = "",
  legend.labs  = c("Low cytotoxic", "High cytotoxic"),
  ggtheme      = theme_hds()
)

# Strip the fill-box from the legend key (leaves only the coloured line)
fig_2a <- km_plot$plot +
  guides(
    colour = guide_legend(override.aes = list(fill = NA)),
    fill   = "none"
  ) +
  scale_y_continuous(
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    labels = c("0.00", "0.25", "0.50", "0.75", "1.00"),
#    limits = c(-0.05, 1.05)
  ) 
# Save the plot as SVG, fiddle with the size here.
ggsave("poster/fig_2a.svg", fig_2a, width = 6, height = 5, device = "svg")
fig_2a
```

#### 2B. I-SPY2 logistic curves

```{r}
#| label: plot-2b-logistic
#| echo: true
#| code-fold: true
#| code-summary: "Figure 2B code"
#| fig-width: 9
#| fig-height: 5

# --- Prepare data: stack pembro + control with arm label and pooled cyto_z --
# `pooled` is in memory from 05-ispy2.R (pembro + control, scaled once)
ispy_plot_data <- pooled %>%
  mutate(
    arm_label = factor(
      ifelse(is_pembro == 1, "Pembrolizumab", "Control"),
      levels = c("Pembrolizumab", "Control")
    )
  )

# --- Build legend labels with n, OR, p from the logistic results CSV --------
ispy_results <- read.csv("data/processed/ispy2_pembro_logistic.csv")
or_pembro  <- ispy_results$OR[grepl("Pembrolizumab", ispy_results$arm)]
p_pembro   <- ispy_results$p_value[grepl("Pembrolizumab", ispy_results$arm)]
n_pembro   <- ispy_results$n[grepl("Pembrolizumab", ispy_results$arm)]
or_control <- ispy_results$OR[grepl("control", ispy_results$arm, ignore.case = TRUE)]
p_control  <- ispy_results$p_value[grepl("control", ispy_results$arm, ignore.case = TRUE)]
n_control  <- ispy_results$n[grepl("control", ispy_results$arm, ignore.case = TRUE)]

# Interaction p-value (cytotoxic × arm)
interaction_df <- read.csv("data/processed/ispy2_interaction.csv")
p_interact <- interaction_df$Pr...z..[grepl("Cytotoxic.*is_pembro", interaction_df$term)]

arm_labels <- c(
  "Pembrolizumab" = sprintf("Pembrolizumab (n = %d, OR = %.2f, p = %.3f)", n_pembro, or_pembro, p_pembro),
  "Control"       = sprintf("Control (n = %d, OR = %.2f, p = %.3f)", n_control, or_control, p_control)
)

fig_2b <- ggplot(ispy_plot_data,
                 aes(x = Cytotoxic_cells_z, y = pcr, colour = arm_label, fill = arm_label)) +
  # Rug marks for individual patients (top for pCR=1, bottom for pCR=0)
  geom_point(
    position = position_jitter(height = 0.03, width = 0),
    alpha = 0.5, size = 1.2
  ) +
  # Logistic regression curve per arm with 95% CI ribbon
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE, alpha = 0.15, linewidth = 0.8
  ) +
  scale_colour_arm(
    labels = arm_labels,
    guide  = guide_legend(override.aes = list(fill = NA))  # strip fill from key
  ) +
  scale_fill_arm(guide = "none") +  # hide fill from legend — only line shows
  scale_y_continuous(
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    labels = c("0.00", "0.25", "0.50", "0.75", "1.00"),
    limits = c(-0.05, 1.05)
  ) +
  annotate("text", x = Inf, y = 0.05,
           label = sprintf("Interaction p = %.2f", p_interact),
           hjust = 1, size = 3, colour = "grey40") +
  labs(
    x = "Cytotoxic immune cell population score (normalised & pooled)",
    y = "Pathological complete response",
    colour = NULL,
    fill = NULL
  ) +
  theme_hds() +
  theme(
    legend.position = "top",
    legend.justification = "left"
  )

# Save the plot as SVG, fiddle with the size here.
ggsave("poster/fig_2b.svg", fig_2b, width = 6, height = 5, device = "svg")
fig_2b
```

**Figure 2** | **Cytotoxic infiltration associates with survival and chemotherapy response.**
**(A)** Kaplan–Meier overall survival in METABRIC by median split of cytotoxic score.
**(B)** Logistic regression of pathological complete response against pooled cytotoxic score in the I-SPY2 pembrolizumab + paclitaxel arm (n = 69) and paclitaxel-only control (n = 179). Both arms show a significant association; the interaction term is non-significant (p = 0.35), suggesting a prognostic rather than predictive effect. Shaded bands: 95% CI.


# Data & methods

## Data sources

- METABRIC — breast cancer cohort (n ≈ 1,904) [@curtis2012genomic]
- TCGA BRCA — PanCancer Atlas 2018 cohort (n ≈ 1,082) [@tcga2012comprehensive; @hoadley2018cell]
- I-SPY2 — neoadjuvant trial microarray cohort (GSE194040) [@wolf2022redefining]
- Danaher panel — immune cell gene markers (14 populations) [@danaher2017gene]

## Data access tools

- cBioPortal — API access for METABRIC and TCGA BRCA [@cerami2012cbioportal; @gao2013integrative; @debruijn2023analysis]
- GEO — public archive hosting I-SPY2 (GSE194040) [@edgar2002geo; @clough2024geo]

## Software

- R statistical computing environment [@r2026]
- survival — Cox proportional-hazards and Kaplan–Meier estimation [@therneau2000survival]
- ggplot2 — grammar-of-graphics visualisation [@wickham2016ggplot2]
- tidyverse — data wrangling (dplyr, tidyr) [@wickham2019tidyverse]
- cBioPortalData — Bioconductor client for cBioPortal API [@ramos2020cbioportaldata]
- GEOquery — Bioconductor client for GEO [@davis2007geoquery]

# References

::: {#refs}
:::


# Supplementary figures

## Cox assumption tests

### S1. Schoenfeld residuals (proportional hazards)


```{r}
#| label: plot-sup-schoenfeld
#| fig-width: 10
#| fig-height: 8

# Manually extract Schoenfeld residual data from cox.zph() objects
# (ph_m = METABRIC f3, ph_t = TCGA g3 from 04-confirmatory-cox.R)
# Canonical display labels so METABRIC and TCGA terms line up in the facets.
# Without this, factor(tumor_stage) (METABRIC) and factor(tumor_stage_num)
# (TCGA) land in different rows of facet_grid().
clean_term <- function(x) {
  dplyr::case_when(
    x == "cyto_z"                  ~ "Cytotoxic (z)",
    x == "pam50_subtype"           ~ "PAM50 subtype",
    x == "age_z"                   ~ "Age (z)",
    grepl("tumor_stage", x)        ~ "Tumour stage",
    x == "GLOBAL"                  ~ "GLOBAL",
    TRUE                           ~ x
  )
}

extract_schoenfeld <- function(ph_obj, cohort_label) {
  terms <- colnames(ph_obj$y)
  pvals <- ph_obj$table[terms, "p"]
  # Long-format data frame: one row per (term, time) residual point
  do.call(rbind, lapply(seq_along(terms), function(i) {
    data.frame(
      cohort     = cohort_label,
      term       = terms[i],
      term_clean = clean_term(terms[i]),
      term_label = sprintf("%s (p=%.3f)", clean_term(terms[i]), pvals[i]),
      time       = ph_obj$x,
      resid      = ph_obj$y[, i],
      stringsAsFactors = FALSE
    )
  }))
}

term_order <- c("Cytotoxic (z)", "PAM50 subtype", "Age (z)", "Tumour stage")

sch_df <- bind_rows(
  extract_schoenfeld(ph_m, "METABRIC"),
  extract_schoenfeld(ph_t, "TCGA")
) %>%
  mutate(
    cohort     = factor(cohort, levels = c("METABRIC", "TCGA")),
    term_clean = factor(term_clean, levels = term_order)
  )

fig_sup_sch <- ggplot(sch_df, aes(x = time, y = resid, colour = cohort, fill = cohort)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = PALETTE$neutral, linewidth = 0.4) +
  geom_point(alpha = 0.3, size = 0.6) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 0.6) +
  facet_grid(term_clean ~ cohort, scales = "free", labeller = label_wrap_gen(width = 15)) +
  scale_colour_cohort(guide = "none") +
  scale_fill_cohort(guide = "none") +
  labs(
    x = "Transformed survival time",
    y = "Scaled Schoenfeld residual"
  ) +
  theme_hds(base_size = 10)

fig_sup_sch
```

### S2. Martingale residuals (linearity)

Martingale residuals from a null Cox model (confounders only, no cytotoxic) plotted against the raw cytotoxic score.

```{r}
#| label: plot-sup-martingale
#| fig-width: 10
#| fig-height: 4

# METABRIC: mart_resid_m from 04, surv_m_stg contains the raw cytotoxic column
mart_df_m <- data.frame(
  cohort    = "METABRIC",
  cytotoxic = surv_m_stg$Cytotoxic_cells,
  resid     = mart_resid_m
)
mart_df_t <- data.frame(
  cohort    = "TCGA",
  cytotoxic = surv_t_stg$Cytotoxic_cells,
  resid     = mart_resid_t
)
mart_df <- bind_rows(mart_df_m, mart_df_t) %>%
  mutate(cohort = factor(cohort, levels = c("METABRIC", "TCGA")))

fig_sup_mart <- ggplot(mart_df, aes(x = cytotoxic, y = resid, colour = cohort, fill = cohort)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = PALETTE$neutral, linewidth = 0.4) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", alpha = 0.15, linewidth = 0.6) +
  facet_wrap(~ cohort, scales = "free_x") +
  scale_colour_cohort(guide = "none") +
  scale_fill_cohort(guide = "none") +
  labs(
    x = "Cytotoxic score (raw)",
    y = "Martingale residual"
  ) +
  theme_hds()

fig_sup_mart
```

## PCA variance structure

### S3. Scree plot

PC1 dominates the immune axis with ~38% variance.

```{r}
#| label: plot-sup-scree
#| fig-width: 6
#| fig-height: 4

# pca_m comes from 03-pca-cytotoxic.R
scree_df <- data.frame(
  pc       = factor(paste0("PC", seq_along(summary(pca_m)$importance[2, ])),
                    levels = paste0("PC", seq_along(summary(pca_m)$importance[2, ]))),
  variance = summary(pca_m)$importance[2, ] * 100
)

fig_sup_scree <- ggplot(scree_df, aes(x = pc, y = variance, group = 1)) +
  geom_hline(
    yintercept = 100 / nrow(scree_df),
    linetype = "dashed", colour = PALETTE$neutral, linewidth = 0.4
  ) +
  geom_line(colour = PALETTE$metabric, linewidth = 0.6) +
  geom_point(colour = PALETTE$metabric, size = 2.5) +
  labs(
    x = NULL,
    y = "Variance explained (%)",
    caption = "Dashed line: equal variance baseline (100% / 14 PCs)"
  ) +
  theme_hds() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig_sup_scree
```
